---
title: "rres_proj1"
date: "November 22, 2016"
output: 
  html_document: 
    keep_md: yes
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(base.dir = 'docs')
```
# Setup - Load Data and Libraries
Load data

```{r load_data, Include=TRUE, }
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
activity.csv <-unzip(temp)
activity <- read.csv(activity.csv, sep=",", header = TRUE, stringsAsFactors = FALSE)
unlink(temp)
```

Load Libraries
```{r libs, include=TRUE, results='hide', message=FALSE, warning=FALSE}
##Load libraries
library(dplyr)
library(lubridate)
library(lattice)
```

# What is mean total number of steps taken per day

#### 1) Calculate the Total number of steps taken per day
```{r tot_steps, include=TRUE}
step1 <- aggregate(steps ~ date, data = activity, sum)
colnames(step1)[2] <- "step_sum"
```

#### 2) Make a histogram of the total number of steps taken each day
```{r hist1, include=TRUE}
range <- length(step1$step_sum)
hist(step1$step_sum, breaks=range, main="Histogram of Total Steps Taken per Day", xlab = "Total Daily Steps", col = "green")
```

#### 3) Calculate and report the mean and median of the total number of steps taken per day
```{r mean_med, include=TRUE}
tstepmean <- mean(step1$step_sum)
tstepmedian <- median(step1$step_sum)
```
mean = `r format(tstepmean, digits = 1L)`

median = `r tstepmedian`

#### 4) Create a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r avesteps, include=TRUE}
avesteps <- aggregate(steps ~ interval, data = activity, mean, na.rm=TRUE)
par(mfrow=c(1,1))
with(avesteps, plot(interval, steps, type = "l", main = "Average Number of Steps by 5min Interval averaged across all days", xlab = "5 min intervals over 24 hours period", ylab = "Number of steps"))
```

#### 5) Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r max_steps, include=TRUE}
b <- which.max(avesteps$steps)
c <- avesteps$interval[b]
d <- avesteps$steps[b]
```

The maximum number of steps were `r format(d, digits = 1L)` 
taken at military time `r c` 

# Imputing missing values

#### 1) Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing, include=FALSE}
e <- complete.cases(activity)
p <- sum(!e) 
```

The total number of rows with missing values is `r p`.

#### 2) Devise a strategy for filling in all of the missing values in the dataset. 

Replacing NA values with mean step count.

#### 3) Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r new_dataset, include=TRUE}
activity2 <- activity 
m1 <- mean(activity$steps, na.rm = TRUE)
activity2$steps[is.na(activity2$steps)] <- m1
```

#### 4) Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r histogram1, include=TRUE}
step2 <- aggregate(steps ~ date, data = activity, sum)
colnames(step2)[2] <- "step_sum"
range <- length(step2$step_sum)
hist(step2$step_sum, breaks=range, main="Histogram of Total Steps Taken per Day \n Imputed Dataset", xlab = "Total Daily Steps", col = "green")
astepmean <- mean(step2$step_sum)
astepmedian <- median(step2$step_sum)
```

mean total steps = `r format(astepmean, digits=1L)`

median median total steps = `r astepmedian`


#### 5) Do these values differ from the estimates from the first part of the assignment? 

No, as expected, using the mean and median to calculate the mean and median in a duplicate dataset should not shift those values.  However, the total number of steps taken will change.


#### 6) Create a difference of the two data sets

```{r totsteps_diff, include=TRUE}

k1 <- sum(activity2$steps) 
k2 <- sum(activity$steps, na.rm=TRUE) 
k3 <- k1 - k2
k3 
```

The difference in steps is `r format(k3, digits=1L)`.

#### 7) What is the impact of imputing missing data on the estimates of the total daily number of steps?

`r format(k3, digits=1L)` steps more in the imputed data from the original data. 

This is a `r format((k3/k2)*100, digits=1L)` percent increase.

As expected, by definition total steps will increase when data is imputed since more data points with positive values will be counted for a positive valued dataset.

# Are there differences in activity patterns between weekdays and weekends?

#### Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r factor, include=TRUE}
activity2[,4] <- NULL
activity2[,4] <- as.character()
colnames(activity2)[4] <- "w_day"
activity2$date <- ymd(activity2$date)
dayw <- c("Saturday", "Sunday")

p <- length(activity2$date)
for(i in 1:p)
    if(weekdays(activity2$date[i]) %in% dayw){activity2$w_day[i] <- "weekend"
    } else {
    activity2$w_day[i] <- "weekday"}
activity2[,4] <- as.factor(activity2[,4])
```

#### Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r panel_plot, include=TRUE}
activity22 <- aggregate(steps ~ interval + w_day, data = activity2, mean)
xyplot(steps~interval|w_day, data=activity22, 
       main="Steps by Interval", xlab="Interval",  ylab="Number of Steps",layout=c(1,2), type="l", lwd=1, col="black")

```

Findings:  weekend steps/intervals are higher after 9:00 than the weekday counts.